---
title: "STA3030F Project1"
author: "Shamiso Chikamhi"
date: "10/03/2020"
output:
  word_document: default
  pdf_document: default
---
Question 1:

```{r results='hide'}
# Defining data
mydata <- read.delim("F:/project 1/incexp.txt")
index = 167
age = mydata$AGE[index:(index+99)] 
hhinc =mydata$HHINCOME[index:(index+29)]
```

Part A
Exploratory data analysis.
```{r results='hide'}
# AGE variable

boxplot(age,main = "The age of the head of the household", col = "gray" )
cat("The summary statistics of the age variable are given as:")
summary(age)
hist(age, breaks = 15, main ="The age of the head of the household", col ="gray", border="blue" )

# House hold income variable
boxplot(hhinc, main = "total monthly income of the household", col = "gray" )
cat("The summary statistics of the household income variable are given as:")
summary(hhinc)
hist(hhinc, main ="total monthly income of the household.", col ="gray", border="blue" )
```

Part B
95% confidence interval for the population mean age.
```{r results='hide'}
sample_mean_age = mean(age)

# Bootstrap procedure for the mean
age_bts <- matrix(0,ncol=100,nrow=4000)
for( i in 1:4000)
{samp=sample(age,size=100,replace=TRUE)
age_bts[i,]=samp}
age_bts=apply(age_bts,1,mean)
age_sort_bts=sort(age_bts)

# Bootstrap confidence interval
cat("The bootstrap 95% confidence interval is (",age_sort_bts[0.025*4000],",",age_sort_bts[0.975*4000],")")

# Lower and upper bound error
lb_error = age_sort_bts[0.025*4000] - sample_mean_age
ub_error = age_sort_bts[0.975*4000] - sample_mean_age
cat("The lower bound error is", lb_error, "and the upper bound error is",ub_error)

#Confidence interval bounds
bound1 = sample_mean_age - lb_error
bound2 = sample_mean_age - ub_error

#95% confidence interval
if (bound1<bound2){
  cat("The 95% confidence interval for the population mean age is (", bound1, ",", bound2, ")")
}  else{   
  cat("The 95% confidence interval for the population mean age is (", bound2, ",", bound1, ")")
}

```

Part C 
Hypothesis test that age is less than or equal to 43
H0, H1
```{r results='hide'}
age_null_hyp = 43
error = sample_mean_age - age_null_hyp
cat("The sample mean is", sample_mean_age, "and the sampling error is", error)

# vector of elements that are greater than smaple mean plus error
pv_elements <- age_bts > (sample_mean_age+error)
cat("The number bootstrap means greater than", sample_mean_age+error, "is:")
sum(pv_elements)
cat("Hence the p-value is:")
p_value = (sum(pv_elements))/4000

# P-value interpretation
if (p_value < 0.05){
  cat ("The p-value is", p_value, "which is significantly small hence we reject null hypothesis and conclude that the mean age is greater than 43.")
} else {
cat ("The p-value is", p_value, "which is significantly large hence we fail to reject null hypothesis and conclude that the mean age is less than 43.")
}
```

Part D 
histogram of the distribution of the sample mean age.
```{r results='hide'}
hist(age_bts, breaks = 20, main ="The bootstrap mean age of the head of the household", col ="gray", border="blue")
```

Part E
90 % confidence interval of the median household income.
```{r results='hide'}
sample_median_hhinc = median(hhinc)
cat("The sample median is", sample_median_hhinc)

# Bootstrap procedeure fro the meadian of household income
hhinc_bts <- matrix(0,ncol=30,nrow=4000)
for( i in 1:4000)
{samp1=sample(hhinc,size=30,replace=TRUE)
hhinc_bts[i,]=samp1}
hhinc_bts=apply(hhinc_bts,1,median)
hhinc_sort_bts=sort(hhinc_bts)

# Bootstrap confidence interval
cat("Bootstrap 90% confidence interval is (",hhinc_sort_bts[0.05*4000],",",hhinc_sort_bts[0.95*4000],")")

# Lower and upper bound error
lb1_error = hhinc_sort_bts[0.025*4000] - sample_median_hhinc
ub1_error = hhinc_sort_bts[0.975*4000] - sample_median_hhinc
cat("The lower bound error is", lb1_error, "and the upper bound error is", ub1_error)

# Confidence Interval bounds 
bound3 = sample_median_hhinc - lb1_error
bound4 = sample_median_hhinc - ub1_error
# 90% confidence interval
if (bound1<bound2){
  cat("The 90% confidence interval for the house hould income median is (", bound3, ",", bound4, ")")
}  else{   
  cat("The 90% confidence interval for the house hould income median is  (", bound3, ",", bound4, ")")
}

```

Question 2
```{r results='hide'}
# Defining data
data2 <-  read.delim("F:/project 1/E1.txt")
fertA = data2$Fert_A
fertB = data2$Fert_B

# Data summary
cat("The data summary of yield from fertiliser A is:")
summary(fertA)
cat("The data summary of yield from fertiliser B is:")
summary(fertB)

```

Part A
Test whether there are significant differences in means between the two groups.
H0 and H1
```{r results='hide'}
# Original sample mean differences
samp_meanA = mean(fertA)
samp_meanB = mean(fertB)
samp_meanDiff = samp_meanB - samp_meanA
cat("The original sample mean difference", samp_meanDiff)

# Bootstrap for the mean differences between the 2 groups
all_A_B =c(fertA,fertB)
bstA=bstB=matrix(0,ncol=20,nrow=5000)
for(j in 1:5000)
{samp2=sample(all_A_B,size=40,replace=TRUE)
bstA[j,] =samp2[1:20]
bstB[j,]=samp2[21:40]
}

bstA_mean=apply(bstA,1,mean)
bstB_mean=apply(bstB,1,mean)
bst_mean_diff=bstB_mean-bstA_mean

# Vector of elements that are greater than smaple mean plus error
pv_diff <- (bst_mean_diff > samp_meanDiff)|(bst_mean_diff < -samp_meanDiff)
cat("The number of bootstrap differences execeding or are less than the sample mean difference:")
sum(pv_diff)
p_value1 = sum(pv_diff)/5000 
cat("The p-value is", p_value1)


# Interpretaion of the p-value
if (p_value1 < 0.05){
  cat ("The p-value is", p_value1, "which is significantly small hence we reject null hypothesis and conclude that the mean crop yields from the two fertilisers are not equal.")
} else {
  cat ("The p-value is", p_value1, "which is significantly large hence we fail to reject null hypothesis and conclude that the mean crop yields from the two fertilisers are equal.")
}

```

Part B
95% confidence interval for the difference in means
```{r results='hide'}
# Bootstrap confidence interval
sort_bst_mean_diff = sort(bst_mean_diff)
cat("The bootstrap difference 95% confidence interval is (", sort_bst_mean_diff[0.025*5000],",",sort_bst_mean_diff[0.975*5000],")")

#Lower and upper bound error
lb2_error = sort_bst_mean_diff[0.025*4000] - samp_meanDiff
ub2_error = sort_bst_mean_diff[0.975*4000] - samp_meanDiff
cat("The lower bound error is", lb2_error, "and the upper bound error is", ub2_error)

# Confidence interval bounds 
bound5 = samp_meanDiff - lb2_error
bound6 = samp_meanDiff - ub2_error
# Confidence interval
if (bound5<bound6){
  cat("The 95% confidence interval for the mean yield difference is (", bound5, ",", bound6, ")")
}  else{   
  cat("The 95% confidence interval for the mean yield difference is (", bound6, ",", bound5, ")")
}
```

Part C
Test for the variances for equality.
```{r results='hide'}
# Original sample F-ratio
varA = var(fertA)
varB = var(fertB)
samp_F = varA/varB
cat("The variance of yield from fetilier A is", varA,"while that of fertiliser B is", varB,". The original sample F-ratio is ", samp_F)

# Bootstrap F-ratio
bstA_var =apply(bstA,1,var)
bstB_var =apply(bstB,1,var)
bst_F = bstA_var/bstB_var

# Vector of elements that are greater than the sample F ratio
pv_F <- (bst_F > samp_F)
cat("The number of bootstrap F-ratio greater than the original F-ratio:")
sum(pv_F)
p_value2 = sum(pv_F)/5000
cat("The p-value is", p_value2)


# Interpretaion of the p-value
if (p_value1 < 0.05){
  cat ("The p-value is", p_value2, "which is significantly small hence we reject null hypothesis and conclude that the variance of the crop yields from the two fertilisers are different.")
} else {
  cat ("The p-value is", p_value2, "which is significantly large hence we fail to reject null hypothesis and conclude that the variane of the crop yields from the two fertilisers are equal.")
}

```

Part D
Comparison of  the results from bootstrapping and normal theory
```{r results='hide'}
# Calculating the p-value
df1 = length(fertA)-1
df2 = length(fertB)-1
p_value3 = pf(samp_F, df1,df2)
cat("An F-test with", df1,",",df2, "Degrees of freedom is conducted and the p-value is", p_value3)

# Interpreting the p-value
if (p_value1 < 0.05){
  cat ("The p-value is", p_value3, "which is significantly small hence we reject null hypothesis and conclude that the variance of the crop yields from the two fertilisers are different.")
} else {
  cat ("The p-value is", p_value3, "which is significantly large hence we fail to reject null hypothesis and conclude that the variane of the crop yields from the two fertilisers are equal.")
}


```

Question 3
```{r results='hide'}
# Defining data
data3 <- read.delim("F:/project 1/aov1.txt")
data3 <- na.omit(data3) 
p1= data3$Price_9
p2 = data3$Price_10
p3 = data3$Price_11

price = list("numeric",length=3)
price[[1]]=c(p1)
price[[2]]=c(p2)
price[[3]]=c(p3)

n=c(length(price[[1]]),length(price[[2]]),length(price[[3]]))
N=sum(n)
k=length(price)

```

Part A 
test whether there are significant differences in means between the groups
H0 H1
```{r results='hide'}
#ANOVA WITH ORIGINAL DATA
RATIOS = Yi.=vector("numeric")
Y.. = mean(unlist(price))
cat("The grand mean is", Y..)
SSE = SST = 0


for(j in 1:k)
{Yi.[j] = mean(price[[j]])
SSE = SSE+sum((price[[j]]-Yi.[j])^2)
SST = SST+n[j]*(Yi.[j]-Y..)^2
}

cat("The mean for price1 is",Yi.[1],"for price 2 is", Yi.[2], "and for price 3 it is", Yi.[3],"The SSE is",SSE, "and the SST is",SST)
F_sample = (SST/(k-1))/(SSE/(N-k))
cat("The original F-ratio is", F_sample)

# ANOVA with bootstrap samples
B3 = 5000
ALLprice = unlist(price)

bst1=bst2=bst3 = matrix(0, nrow = 5000, ncol = 20 )
for(h in 1:B3)
{SSE = SST = 0 
Yi.=vector("numeric")
price = list("numeric",length=3)
bst_anova = sample(ALLprice, size=60, replace=TRUE)
bst1[h,] = bst_anova[1:20]
bst2[h,] = bst_anova[21:40]
bst3[h,] = bst_anova[41:60]

price[[1]]=bst_anova[1:20]
price[[2]]=bst_anova[21:40]
price[[3]]=bst_anova[41:60]

for(j in 1:k)
{Yi.[j] = mean(price[[j]])
SSE = SSE+sum((price[[j]]-Yi.[j])^2)
SST = SST+n[j]*(Yi.[j]-Y..)^2
}
RATIOS[h]=(SST/(k-1))/(SSE/(N-k))
}
```
 
Part B
print-out of the first 3 bootstrap samples generated
```{r results='hide'}
# First 3 bootstrap samples
cat("Samples for price 1:")
bst1[1:3,]
cat("Samples for price 2:")
bst2[1:3,]
cat("Samples for price 3:")
bst3[1:3,]


```

Part C
p-value and conclusion based on the bootstrap results.
```{r results='hide'}
p_value4 = length(RATIOS[RATIOS>F_sample])/B3

theory_pvalue = pf(F_sample, k-1, N-k, lower.tail = FALSE)
cat("The bootstrap pvalue is", p_value4)
cat("The theoretical pvalue is", theory_pvalue)


# Interpretation of the p-value
if (p_value1 < 0.05){
  cat ("The p-value is", p_value4, "which is significantly small hence we reject null hypothesis and conclude that there are significant differences in means between the groups.")
} else {
  cat ("The p-value is", p_value4, "which is significantly large hence we fail to reject null hypothesis and conclude that there are no significant differences in means between the groups.")
}

```

Part D
Interpretation of the results.
